Long- and short-read metabarcoding technologies reveal similar spatio-temporal structures in fungal communities

### 2020-09-30 writing/transect_paper.Rmd
add accumulation plots
work on reviewer comments

Different sequencing technologies have different capabilities in terms of sequencing depth and read length, as well as differing quality profiles and potential biases [@yang2013].
The rapid development of new HTS technologies, as well as subsequent iterative improvements in sequencing chemistry and read capacity, means that the technologies used in metabarcoding studies, along with any associated biases, change rapidly.
As an example, the first study using HTS metabarcoding of soil fungi was published in 2009 [@buee2009] using 454 Pyrosequencing; production of 454 sequencers was subsequently discontinued in 2015, and sales of reagents stopped in 2016 [@hollmer2013].
This brings into question the comparability of studies conducted only a few years apart.

@@

Sampling was conducted at two sites (Ang: N 9.75456° E 2.14064°; Gan: N 9.75678° E 2.31058°) approximately 30\ km apart in the _Forêt Classée de l'Ouémé Supérieur_ (Upper Ouémé Forest Reserve) in central Benin.
Both sites were located in Soudanian savannah woodlands [@olson2001;@yorou2014] dominated by the ECM host tree _Isoberlinia doka_ (Caesalpinioideae).
At each site, 25\ soil samples were collected along a single 24 m linear transect at intervals of 1\ m in May 2015.
One third of the sample locations (3\ m spacing) were resampled one year later in June 2016.
For each sample, any coarse organic debris <!--Do you have any definition? Wider than 5 mm? MR-->was removed from the soil surface and a sample of approximately 5cm×5cm×5cm was extracted with a sterilized knife blade.
Each sample was sealed in a plastic zipper bag and homogenized by shaking and manually breaking apart soil aggregations.
Approximately 250\ mg total of soil was collected from two locations in the homogenized soil sample and placed into a separate 2.0\ mL microtube containing 750\ mL of field lysis/preservation buffer and lysis beads (Xpedition^TM^ Soil/Fecal DNA miniprep, Zymo Research Corporation, Irvine, California, USA).
Due to the inavailability of refrigeration, extraction was initiated in the field according to the manufacturers instructions by lysis using a handheld bead-beater (TerraLyser^TM^; Zymo Research Corporation).

An additional sample was collected at every sampling location (1-m spacing) in 2016 using LifeGuard^TM^ Soil Preservation Solution (MO BIO, Carlsbad, CA; USA) for preservation, without field lysis.
Sequencing results for these samples differed significantly (PERMANOVA with 9999 permutations, $p < `r readd(buffer_adonis)["buffer","p.value"] %>% formatC(format = "f")`$, $R^2 = `r readd(buffer_adonis)["buffer","R2"] %>% formatC(format = "f", digits = 2)`$) from samples preserved using the Xpedition^TM^ lysis buffer (Figures \@ref(fig:buffer-pcoa), \@ref(fig:buffer-compare-long), and \@ref(fig:buffer-compare-short)); as such the LifeGuard preserved samples were excluded from our spatial analyses.
However, reads from these samples were included in the full bioinformatics workflow, including ASV calling, OTU clustering, and phylogenetic trees.

@@

After pooling of technical replicates, amplicons were purified using SPRI beads [@vesterinen2016] and quantified fluorometrically as above.
An aliquot of 100\ ng of DNA from each sample (or the total PCR product if less than 100\ ng) was pooled into two libraries each for long and short amplicons.
Each library was sequenced using Single Molecule Real Time (SMRT) sequencing on a Pacific Biosciences (PacBio) RS II sequencer at the Uppsala Genome Center (UGC; Uppsala Genome Center, Science
for Life Laboratory, Dept. of Immunology, Genetics and Pathology, Uppsala University, BMC, Box 815, SE-752\ 37 UPPSALA, Sweden).
Short amplicon libraries were sequenced on two SMRT cells each, while long amplicon libraries were sequenced on four SMRT cells each.

Additionally, the same short amplicon PCR libraries were combined and sequenced using an Ion S5 (Ion Torrent) sequencer using one 520 chip at UGC, and a MiSeq (Illumina Inc.) sequencer using v3 chemistry with a paired-end read length of 300\ bp at the SNP&SEQ Technology Platform (Dept. of Medical Sciences, Uppsala University, BMC, Box 1432, SE-751\ 44 UPPSALA, Sweden) using one half of a lane.

@@

Circular consensus sequence (CCS) basecalls for PacBio sequences were made using `ccs` version 3.4 [@pacificbiosciences2019] using the default settings.
The resulting sequences, as well as the paired-end Illumina sequences, were demultiplexed and sequencing primers were removed using `cutadapt` version 2.8 [@martin2011].
Sequencing primers were similarly removed from the Ion Torrent sequences, but interference between the tagged gITS7 primers and the Ion XPress tags used in library prep made full demultiplexing of the Ion Torrent sequences impossible, and these reads were thus only analyzed as a pool.
For Ion Torrent and PacBio, reads were discarded if they did not have the appropriate primers on both ends.
Reads were searched in both directions, and reads where the primers were found in the reverse direction were reverse complemented before further analysis.
For Illumina sequences, read pairs were only retained when PCR primers were detected at the 5' ends of both the forward and reverse read.
Primers were also searched for and removed on the 3' ends of the reads, in case of readthrough with short amplicons.
Read pairs where the primers were found in reverse orientation were kept in separate files, but were retained in their original orientation until after denoising.

@@

<!-- R3: honestly this section is above my head and I can’t judge its accuracy. Hopefully another review is more bioinformatically inclined and can judge it better. -->
Raw reads were divided into shorter regions by matching to covariance models (CM), which are similar to stochastic hidden markov models (HMM), but account for both nucleotide sequence and RNA secondary structure [@eddy1994].
First, the 5.8S rDNA was located in each read by searching for Rfam model RF0002 [@kalvari2018] using `cmsearch` from Infernal 1.1.2 [@nawrocki2013], and all bases before the 5.8S were assigned to ITS1.
No attempt was made to remove the approximately 12\ bp fragment of the SSU from the 5' end of ITS1 in the long amplicons; it was too short to be reliably detected by a CM or the HMMs employed by ITSx [@bengtsson-palme2013].
<!--This approach was able to identify both the 5' and 3' ends of the 5.8S rDNA more consistently and quickly than ITSx [@bengtsson-palme2013, data not shown].-->
A reference alignment including conserved RNA base pairing between and within the 5.8S and relevant portions of LSU was generated from the fungal 28S RNA seed alignment from RDP release 11.5 [@glockner2017; @cole2014] by truncating after the LR5 primer site and using the reference line to annotate the variable regions *sensu* @michot1984 and @raue1988.
A CM was generated from the alignment using `cmbuild` from Infernal.
The fragment of each read beginning with the 5.8S rDNA was then aligned to the CM using `cmalign` from Infernal.
The annotation line in the CM alignment for each read was then used to split the reads into alternating more-conserved and less-conserved domains as shown in Figure\ \@ref(fig:regions),
where LSU1-4 represent the conserved domains of LSU flanking the variable D1-3 domains [@michot1984].
For short amplicons, only (partial) 5.8S, ITS2, and (partial) LSU1 were extracted.
Code to extract the segments, including annotated seed alignments and CMs, is available in the new R package `LSUx`, available on Github at https://github.com/brendanf/LSUx.

@@

ASVs from the short amplicon datasets were refined using only the final strict consensus step, i.e., an assignment at a given rank was accepted if there was no conflict between the different assignment algorithms at greater than 50% confidence.
This refinement method is referred to as "Consensus".
Additionally, a hybrid method, was applied to the short amplicon datasets, in which assignments from PHYLOTAX were used for ASVs which could be linked by an identical ITS2 region to a long amplicon, and assignments from Consensus were used for the remaining ASVs.

@@

The effect of sequencing technology and amplicon length on the recovered fungal community composition, as assessed by the Bray-Curtis dissimilarity, was determined by PERMANOVA [@anderson2003] using the `adonis2` function in the R package `vegan` [@R-vegan].
Comparisons were made between the fungal communities recovered by the three sequencing strategies that were successfully demultiplexed (Illumina short, PacBio short, PacBio long).
<!-- R3: I would say, the effect of the sequencing strategies and read length on the fungal community compositional dissimilarity, as assessed by either UNIFRAC or Bray-Curtis, was determined by PERMANOVA (cite Marti Anderson) using the Adonis function in vegan. I’m confused because you definitely mentioned in the intro UNIfrac but in this section only mention Bray-Curtis. Also you don’t mention any sort of normalization for differing sequence reads per sample, and that’s not really acceptable. I recommend avg.dist function in vegan. It lets you make the Bray-Curtis dissimilarity matrix subsampling for an even sequencing depth 100x and then take the median and perform a square root transformation all in one line of code.

Also, what no alpha diversity analysis? Beta-diversity analyses are generally pretty robust. I would be way more interested to see differences in alpha diversity by sequencing technology. Can you please report this (normalized for differences in sequencing depth)?
-->
In order to detect bias at larger taxonomic scales, ASVs were clustered according to the assigned taxonomic class.
Only samples where all three strategies yielded at least 100\ fungal reads (`r phyloseq::nsamples(readd(tech_class_table))/3` samples), and classes which represented at least 1% of reads in at least one sample (`r phyloseq::ntaxa(readd(tech_class_table))` classes), were included.
The community matrix was normalized to proportional abundance by dividing the counts for each class in each sample by the total number of reads for that sample, but were not rarefied to even sampling depth [@mcmurdie2014].
PERMANOVA included three terms: an indicator for soil sample, comprising all spatiotemporal effects; amplicon length (long vs. short); and sequencing technology (Illumina MiSeq vs. PacBio RS II).
Partial Principal Coordinates Analysis (PPCoA) was applied to the same dissimilarity matrix using the `capscale` function in `vegan` [@R-vegan].
Spatiotemporal effects were partialled out in order to visualize effects due to sequencing technology and amplicon length.

@@

DNA concentrations after extraction and PCR, as well as sequencing reads for PacBio and Illumina, are shown per sample in Figure \@ref(fig:sample-depth).
Samples from Ang in 2015 yielded low quantities of DNA, poor PCR performance, and ultimately very few sequencing reads, especially in the long amplicon library, where only one sample produced more than 100\ reads.
Consequently, Ang samples were excluded from spatial analysis, although they were retained for denoising, phylogenetic reconstruction, and taxonomic assignment.

The number of sequencing reads and ASVs at each stage in the bioinformatics pipeline are shown in Table\ \@ref(tab:bioinfo).
Sequencing with PacBio RS II yielded more than twice as many raw reads for long amplicons as for short amplicons, with approximately 125 thousand and 50 thousand reads, respectively.
Ion Torrent Ion S5 and Illumina MiSeq yielded substantially more reads, with 20.7 million and 10.8 million, respectively.
<!-- R3: 20.7 M and 10.8M reads for only 25 samples is a lot of read depth! In the future you can go way lower than that to save money. I would multiplex in hundreds of samples for 20.8 M reads.-->
Demultiplexing, primer trimming, and quality filtering reduced these totals by `r (1-readd(filtercounts_full)/readd(rawcounts))["pb_500"] %>% percent(format = "d")` for PacBio long amplicons, but only by `r (1-readd(filtercounts_full)/readd(rawcounts))["pb_483"] %>% percent(format = "d")` for PacBio short amplicons, resulting in a similar number of filtered reads for the two strategies.
Losses in demultiplexing, trimming, and quality filtering were intermediate for Ion Torrent and Illumina, with `r (1-readd(filtercounts_full)/readd(rawcounts))[c("is_057", "SH-2257")] %>% percent(format = "d") %>% text_list()` loss, respectively.
In contrast, extraction of only the ITS2 region before quality filtering resulted in the loss of `r percent(1-readd(filtercounts_ITS2)['pb_500']/readd(rawcounts)['pb_500'], format = "d")` of trimmed long amplicon PacBio reads, `r percent(1-readd(filtercounts_ITS2)['pb_483']/readd(rawcounts)['pb_483'], format = "d")` of trimmed short amplicon PacBio reads, and `r percent(1-readd(filtercounts_ITS2)['is_057']/readd(rawcounts)['is_057'], format = "d")` of trimmed Ion Torrent reads.
This represented greater loss of PacBio short amplicon reads, but less loss of PacBio long amplicon reads and Ion Torrent reads.

Almost all of the short amplicons from all three technologies were between 240 and 375\ bp long (Figure\ \@ref(fig:full-length)a).
Although the length profile of the three sequencing runs were similar, Illumina MiSeq had the largest fraction of reads near the top of the range, followed by Ion Torrent Ion S5 and PacBio RS II (Figure\ \@ref(fig:full-length)b).
The difference in length distributions was statistically significant due to the large sample size (Kruskal-Wallis statistic = `r readd(length_kw)$statistic %>% formatC(format = "e", digits = 2)`, $p < 2.2\times10^{-16}$),
<!-- S3: 2 decimal points is sufficient-->
but the difference between means was fairly small, with mean amplicon lengths of `r filter(readd(reads_table), region == "short") %>% group_by(seq_run) %>% summarize(length = weighted.mean(length, reads)) %>% deframe() %>% magrittr::extract(c("pb_483", "is_057", "SH-2257")) %>% text_list(digits = 0)` bp for PacBio, Ion Torrent, and Illumina, respectively.
In contrast, the length of the long amplicon reads varied widely, from `r filter(readd(reads_table), region == "long") %$% range(length) %>% paste(collapse = " to ")` bp, with a mean of `r filter(readd(reads_table), region == "long") %$% weighted.mean(length, reads) %>% round()` bp.

@@

*A. bisporus* sequences represented `r readd(pos_control_data) %>% filter(complete.cases(.)) %>% group_by(seq_run) %>% summarize(read_frac = sum(pc_reads) / sum(all_reads)) %>% deframe() %>% magrittr::extract(c("pb_500", "pb_483", "SH-2257", "is_057")) %>% percent(format = "f", digits = 2) %>% text_list()` of non-control reads, in the PacBio long, PacBio short, Illumina, and Ion Torrent datasets, respectively, giving similar estimates for the rate of tag-switching for all technologies.
These reads were excluded from community analyses.
<!-- S3: so to clarify, everything was PCR amplified with a primer that had barcodes and adaptor attached to either forward or reverse primer? I was having trouble following this during you PCR description section.-->

@@

(ref:venn-s) Venn diagrams of shared ASVs and OTUs between different sequencing technologies

(ref:venn) Venn diagrams showing shared ITS2-based ASVs (*a*, *c*) and 97% OTUs (*b*, *d*) between different sequencing technologies from the same short amplicon library (*a*, *b*), and between long and short amplicon libraries (*c*, *d*).
In each region, the number of ASVs/OTUs is given above, while the fractions of reads for each sequencing strategy are shown below.
For short amplicons in *c* and *d*, ASV/OTU counts reflect detection by any of the three technologies, and read counts represent the mean fraction of reads across the three technologies.
<!-- S2: The values for Ion Torrent are 1-3 orders of magnitude higher than other methods. This to me is an important finding that is not mentioned(!) -->
<!-- s3: I recommend using “VennDiagram” package in R. It allows you to make Venn diagrams with circle sizes that change based on how many samples it represents. That is way easier to interpret than one with circles that are the same size regardless. I also don’t really understand what the decimals represent – what do you mean by the fraction of reads for each sequencing technology? For example in figure C, is 601 sequences 79% of the long amplicon reads? Please clarify. -->

@@

(ref:read-compare-s) Comparison between read numbers for different sequencing strategies

(ref:read-compare) by ASV (**a**) and 97% OTU (**b**). ASVs/OTUs which were detected by one sequencing strategy but not the other are plotted as tick marks along the axes. Dashed line represents a constant ratio of read numbers. The blue line is a LOESS smooth of the data, with associated uncertainty in grey shading. $R^2$ value displayed is for log-transformed non-zero read numbers.
<!-- R2: Please provide axis labels e.g. panel (a) y-axis is "Count of ASVs" (?) -->
<!-- R3: I’m not sure why this is useful, and I don’t think it warrants so much space in the document. It can be moved to the supplements. To me, a correlation of species richness by read technology might be interesting, or if you had some sort of treatment effect you were looking for and all of the methods gave you the same inferences of treatment effects, then that would be more interesting. I would prefer to see this replaced by correlation of alpha diversity.

I also think species accumulation curves between sequencing technologies would be interesting to see. -->

```{r read-compare, fig.cap="(ref:read-compare-s), (ref:read-compare)", fig.scap="(ref:read-compare-s)", fig.height=8}
ggpubr::ggarrange(
  read_comparison(
    readd(multi_table),
    readd(comparisons),
    type = "ASV"
  ),
  read_comparison(
    readd(multi_table),
    readd(comparisons),
    type = "OTU"
  ),
  ncol = 1,
  labels = "auto",
  label.x = 0.9
)
```

The majority of abundant ASVs and OTUs were captured by all sequencing strategies used (Figure\ \@ref(fig:venn)).
ASVs shared between all datasets represented `r paste(range(as.numeric(unlist(readd(venn_ASV)["1111", c("reads_frac_pb_500", "reads_frac_pb_483", "reads_frac_SH-2257", "reads_frac_is_057")]))*100), collapse = "--")`% of the reads for the long and short PacBio datasets, Illumina dataset, and Ion Torrent dataset, respectively.
These fractions increased to `r paste(range(as.numeric(unlist(readd(venn_OTU)["1111", c("reads_frac_pb_500", "reads_frac_pb_483", "reads_frac_SH-2257", "reads_frac_is_057")]))*100), collapse = "--")`% when differences at the intra-species scale were removed by clustering the ASVs into 97% OTUs.
In particular, `r text_list(percent(colSums(mutate_all(readd(venn_OTU)[c("0111", "1111"), c("reads_frac_pb_483", "reads_frac_SH-2257", "reads_frac_is_057")], as.numeric), na.rm = TRUE)))` of reads in the PacBio, Illumina, and Ion Torrent short-amplicon datasets belonged to OTUs shared between all three datasets.
In contrast, `r percent(readd(venn_ASV)["1000", "reads_frac_pb_500"])` of reads in the long PacBio dataset belonged to ASVs which were unique to that dataset, and the fraction only reduced to `r percent(readd(venn_OTU)["1000", "reads_frac_pb_500"])` after OTU clustering.
<!-- R1: Is this due to the much greater length causing a different ASV than a shorter ASV, even if they have an identical sequence over the same region? That seems to be the case, as the clustering did not really lower the difference between the unique long and short read ASVs at all (going from 21% to 20%).  More depth of explanation in the discussion of this issue, maybe with a couple of examples of what is driving this, would be helpful.-->
Complete tabulations of the number of ASVs and OTUs shared between the different sequencing strategies are shown in Supplementary Tables \@ref(tab:venn-table-asv) and \@ref(tab:venn-table-otu), respectively.

@@

@@

For all sequencing datasets and taxonomic assignment protocols, a higher proportion of reads was assigned than of ASVs, indicating that common ASVs were more likely to be identified than rare ASVs (Figure\ \@ref(fig:tax-chart)).
<!-- R3: do you mean taxonomically identified? -->
A greater fraction of ITS reads and ASVs were assigned using the Unite database than the Warcup database across sequencing technologies, amplicons, algorithms, and taxonomic ranks.
At most taxonomic ranks, the RDPC algorithm assigned the greatest fraction of reads and ASVs, followed by SINTAX, and then IDTAXA.

>>>>

Different sequencing technologies have different capabilities in terms of sequencing depth and read length, as well as differing quality profiles and potential biases [@yang2013].
The rapid development of new HTS technologies, as well as subsequent iterative improvements in sequencing chemistry and read capacity, means that the technologies used in metabarcoding studies, along with any associated biases, change frequently.
As an example, the first study using HTS metabarcoding of soil fungi was published in 2009 [@buee2009] using 454 Pyrosequencing; production of 454 sequencers was subsequently discontinued in 2015, and sales of reagents stopped in 2016 [@hollmer2013].
This brings into question the comparability of studies conducted only a few years apart.

@@

Sampling was conducted at two sites (Ang: N 9.75456° E 2.14064°; Gan: N 9.75678° E 2.31058°) approximately 30\ km apart in the _Forêt Classée de l'Ouémé Supérieur_ (Upper Ouémé Forest Reserve) in central Benin.
Both sites were located in Soudanian savannah woodlands [@olson2001;@yorou2014] dominated by the ECM host tree _Isoberlinia doka_ (Caesalpinioideae).
At each site, 25\ soil samples were collected along a single 24 m linear transect at intervals of 1\ m in May 2015.
One third of the sample locations (3\ m spacing) were resampled one year later in June 2016, for a total of 67 samples.
For each sample, any coarse organic debris <!--Do you have any definition? Wider than 5 mm? MR-->was removed from the soil surface and a sample of approximately 5cm×5cm×5cm was extracted with a sterilized knife blade.
Each sample was sealed in a plastic zipper bag and homogenized by shaking and manually breaking apart soil aggregations.
Approximately 250\ mg total of soil was collected from two locations in the homogenized soil sample and placed into a separate 2.0\ mL microtube containing 750\ mL of field lysis/preservation buffer and lysis beads (Xpedition^TM^ Soil/Fecal DNA miniprep, Zymo Research Corporation, Irvine, California, USA).
Due to the inavailability of refrigeration, extraction was initiated in the field according to the manufacturers instructions by lysis using a handheld bead-beater (TerraLyser^TM^; Zymo Research Corporation).

An additional sample was collected at every sampling location (1-m spacing) in 2016 using LifeGuard^TM^ Soil Preservation Solution (MO BIO, Carlsbad, CA; USA) for preservation, without field lysis.
Sequencing results for these 50 samples differed significantly (PERMANOVA with 9999 permutations, $p < `r readd(buffer_adonis)["buffer","p.value"] %>% formatC(format = "f")`$, $R^2 = `r readd(buffer_adonis)["buffer","R2"] %>% formatC(format = "f", digits = 2)`$) from samples preserved using the Xpedition^TM^ lysis buffer (Figures \@ref(fig:buffer-pcoa), \@ref(fig:buffer-compare-long), and \@ref(fig:buffer-compare-short)); as such the LifeGuard preserved samples were excluded from our spatial analyses.
However, reads from these samples were included in the full bioinformatics workflow, including ASV calling, OTU clustering, and phylogenetic trees.

@@

After pooling of technical replicates, amplicons were purified using SPRI beads [@vesterinen2016] and quantified fluorometrically as above.
An aliquot of 100\ ng of DNA from each sample (or the total PCR product if less than 100\ ng) was pooled into two libraries each for long and short amplicons.

Each library was sequenced using Single Molecule Real Time (SMRT) sequencing on a Pacific Biosciences (PacBio) RS II sequencer at the Uppsala Genome Center (UGC; Uppsala Genome Center, Science
for Life Laboratory, Dept. of Immunology, Genetics and Pathology, Uppsala University, BMC, Box 815, SE-752\ 37 UPPSALA, Sweden).
Short amplicon libraries were sequenced on two SMRT cells each, while long amplicon libraries were sequenced on four SMRT cells each.

Additionally, the same short amplicon PCR libraries were combined and sequenced using an Ion S5 (Ion Torrent) sequencer using one 520 chip at UGC, and a MiSeq (Illumina Inc.) sequencer using v3 chemistry with a paired-end read length of 300\ bp at the SNP&SEQ Technology Platform (Dept. of Medical Sciences, Uppsala University, BMC, Box 1432, SE-751\ 44 UPPSALA, Sweden) using one half of a lane.
Platform-specific library preparation, including adapter ligation, was performed at the sequencing facilities according to their standard protocols.

@@

Circular consensus sequence (CCS) basecalls for PacBio sequences were made using `ccs` version 3.4 [@pacificbiosciences2019] using the default settings.
The resulting sequences, as well as the paired-end Illumina sequences, were demultiplexed and sequencing primers were removed using `cutadapt` version 2.8 [@martin2011].
Sequencing primers were similarly removed from the Ion Torrent sequences, but interference between the tagged gITS7 primers and the Ion XPress tags used in library prep made full demultiplexing of the Ion Torrent sequences impossible, resulting in two samples sharing each tag.
These reads were thus either analyzed as a pool, or comparisons were made to equivalently combined samples in the other datasets.
For Ion Torrent and PacBio, reads were discarded if they did not have the appropriate primers on both ends.
Reads were searched in both directions, and reads where the primers were found in the reverse direction were reverse complemented before further analysis.
For Illumina sequences, read pairs were only retained when PCR primers were detected at the 5' ends of both the forward and reverse read.
Primers were also searched for and removed on the 3' ends of the reads, in case of readthrough with short amplicons.
Read pairs where the primers were found in reverse orientation were kept in separate files, but were retained in their original orientation until after denoising.

@@

Raw reads were divided into shorter regions by matching to covariance models (CM), which are similar to stochastic hidden markov models (HMM), but account for both nucleotide sequence and RNA secondary structure [@eddy1994].
First, the 5.8S rDNA was located in each read by searching for Rfam model RF0002 [@kalvari2018] using `cmsearch` from Infernal 1.1.2 [@nawrocki2013], and all bases before the 5.8S were assigned to ITS1.
No attempt was made to remove the approximately 12\ bp fragment of the SSU from the 5' end of ITS1 in the long amplicons; it was too short to be reliably detected by a CM or the HMMs employed by ITSx [@bengtsson-palme2013].
<!--This approach was able to identify both the 5' and 3' ends of the 5.8S rDNA more consistently and quickly than ITSx [@bengtsson-palme2013, data not shown].-->
A reference alignment including conserved RNA base pairing between and within the 5.8S and relevant portions of LSU was generated from the fungal 28S RNA seed alignment from RDP release 11.5 [@glockner2017; @cole2014] by truncating after the LR5 primer site and using the reference line to annotate the variable regions *sensu* @michot1984 and @raue1988.
A CM was generated from the alignment using `cmbuild` from Infernal.
The fragment of each read beginning with the 5.8S rDNA was then aligned to the CM using `cmalign` from Infernal.
The annotation line in the CM alignment for each read was then used to split the reads into alternating more-conserved and less-conserved domains as shown in Figure\ \@ref(fig:regions),
where LSU1-4 represent the conserved domains of LSU flanking the variable D1-3 domains [@michot1984].
For short amplicons, only (partial) 5.8S, ITS2, and (partial) LSU1 were extracted.
Code to extract the segments, including annotated seed alignments and CMs, is available in the new R package `LSUx`, available on Github at https://github.com/brendanf/LSUx.

@@

ASVs from the short amplicon datasets were refined using only the final strict consensus step, i.e., an assignment at a given rank was accepted if there was no conflict between the different assignment algorithms at greater than 50% confidence.
This refinement method is referred to as "Consensus".
Additionally, a hybrid method was applied to the short amplicon datasets, in which assignments from PHYLOTAX were used for ASVs which could be linked by an identical ITS2 region to a long amplicon, and assignments from Consensus were used for the remaining ASVs.

@@

To assess the completeness of sequencing using each of the sequencing strategies, ASV and OTU accumulation curves were generated using the iNEXT package [citation]. Additionally, average estimated ASV and OTU richness was calculated for each sample after rarefying to 100 sequences using 100 repetitions. ASV and OTU richness were then compared for each biological sample as sequenced by the different strategies. For each pair of strategies, the best fit slope for the ASV richness recovered by each strategy was calculated by total least squares using the "deming" package [citation]. For this analysis, samples in the PacBio and Illumina datasets were combined bioinformatically to match the corresponding samples from the incompletely demultiplexed Ion Torrent dataset.

The effect of sequencing technology and amplicon length on the recovered fungal community composition, as assessed by the Bray-Curtis dissimilarity, was determined by PERMANOVA [@anderson2003] using the `adonis2` function in the R package `vegan` [@R-vegan].
Comparisons were made between the fungal communities recovered by the three sequencing strategies that were successfully demultiplexed (Illumina short, PacBio short, PacBio long).

In order to detect bias at larger taxonomic scales, ASVs were clustered according to the assigned taxonomic class.
Only samples where all three strategies yielded at least 100\ fungal reads (`r phyloseq::nsamples(readd(tech_class_table))/3` samples), and classes which represented at least 1% of reads in at least one sample (`r phyloseq::ntaxa(readd(tech_class_table))` classes), were included.
The community matrix was normalized to proportional abundance by dividing the counts for each class in each sample by the total number of reads for that sample, but were not rarefied to even sampling depth [@mcmurdie2014].
PERMANOVA included three terms: an indicator for soil sample, comprising all spatiotemporal effects; amplicon length (long vs. short); and sequencing technology (Illumina MiSeq vs. PacBio RS II).
Partial Principal Coordinates Analysis (PPCoA) was applied to the same dissimilarity matrix using the `capscale` function in `vegan` [@R-vegan].
Spatiotemporal effects were partialled out in order to visualize effects due to sequencing technology and amplicon length.

@@

Samples from Ang in 2015 yielded low quantities of DNA, poor PCR performance, and ultimately very few sequencing reads, especially in the long amplicon library, where only one sample produced more than 100\ reads (Figure \@ref(fig:sample-depth)).
Consequently, Ang samples were excluded from spatial analysis, although they were retained for denoising, phylogenetic reconstruction, and taxonomic assignment.

The number of sequencing reads and ASVs at each stage in the bioinformatics pipeline differed between sequencing strategies (Table\ \@ref(tab:bioinfo)).
Sequencing with PacBio RS II yielded more than twice as many raw reads for long amplicons as for short amplicons, with approximately 125 thousand and 50 thousand reads, respectively.
Ion Torrent Ion S5 and Illumina MiSeq yielded substantially more reads, with 20.7 million and 10.8 million, respectively.
Demultiplexing, primer trimming, and quality filtering reduced these totals by `r (1-readd(filtercounts_full)/readd(rawcounts))["pb_500"] %>% percent(format = "d")` for PacBio long amplicons, but only by `r (1-readd(filtercounts_full)/readd(rawcounts))["pb_483"] %>% percent(format = "d")` for PacBio short amplicons, resulting in a similar number of filtered reads for the two strategies.
Losses in demultiplexing, trimming, and quality filtering were intermediate for Ion Torrent and Illumina, with `r (1-readd(filtercounts_full)/readd(rawcounts))[c("is_057", "SH-2257")] %>% percent(format = "d") %>% text_list()` loss, respectively.
In contrast, extraction of only the ITS2 region before quality filtering resulted in the loss of `r percent(1-readd(filtercounts_ITS2)['pb_500']/readd(rawcounts)['pb_500'], format = "d")` of trimmed long amplicon PacBio reads, `r percent(1-readd(filtercounts_ITS2)['pb_483']/readd(rawcounts)['pb_483'], format = "d")` of trimmed short amplicon PacBio reads, and `r percent(1-readd(filtercounts_ITS2)['is_057']/readd(rawcounts)['is_057'], format = "d")` of trimmed Ion Torrent reads.
This represented greater loss of PacBio short amplicon reads, but less loss of PacBio long amplicon reads and Ion Torrent reads.

Almost all of the short amplicons from all three technologies were between 240 and 375\ bp long (Figure\ \@ref(fig:full-length)a).
Although the length profile of the three sequencing runs were similar, Illumina MiSeq had the largest fraction of reads near the top of the range, followed by Ion Torrent Ion S5 and PacBio RS II (Figure\ \@ref(fig:full-length)b).
The difference in length distributions was statistically significant due to the large sample size (Kruskal-Wallis statistic = `r readd(length_kw)$statistic %>% formatC(format = "e", digits = 2)`, $p < 2.2\times10^{-16}$),
but the difference between means was fairly small, with mean amplicon lengths of `r filter(readd(reads_table), region == "short") %>% group_by(seq_run) %>% summarize(length = weighted.mean(length, reads)) %>% deframe() %>% magrittr::extract(c("pb_483", "is_057", "SH-2257")) %>% text_list(digits = 0)` bp for PacBio, Ion Torrent, and Illumina, respectively.
In contrast, the length of the long amplicon reads varied widely, from `r filter(readd(reads_table), region == "long") %$% range(length) %>% paste(collapse = " to ")` bp, with a mean of `r filter(readd(reads_table), region == "long") %$% weighted.mean(length, reads) %>% round()` bp.

@@

*A. bisporus* sequences represented `r readd(pos_control_data) %>% filter(complete.cases(.)) %>% group_by(seq_run) %>% summarize(read_frac = sum(pc_reads) / sum(all_reads)) %>% deframe() %>% magrittr::extract(c("pb_500", "pb_483", "SH-2257", "is_057")) %>% percent(format = "f", digits = 2) %>% text_list()` of non-control reads, in the PacBio long, PacBio short, Illumina, and Ion Torrent datasets, respectively, giving similar estimates for the rate of tag-switching for all technologies.
These reads were excluded from community analyses.

@@

(ref:venn-s) Shared richness and abundance of ASVs and OTUs between different sequencing technologies

(ref:venn) Shared richness and abundance of ITS2-based ASVs (*a*, *c*) and 97% OTUs (*b*, *d*) between different sequencing technologies from the same short amplicon library (*a*, *b*), and between long and short amplicon libraries (*c*, *d*).
In each region, the ASV/OTU richness is given above, while the relative abundance of reads represented by these ASVs/OTUs in each sequencing strategy are shown below in the order PacBio/Illumina/Ion Torrent (*a*, *b*), or long/short (*c*, *d*).
For short amplicons in *c* and *d*, ASV/OTU counts reflect detection by any of the three technologies, and read counts represent the mean fraction of reads across the three technologies.

@@

We compared the unique ASVs and OTUs shared between datasets from different sequencing strategies, and the number of reads represented by these ASVs and OTUs in each strategy.
The majority of abundant ASVs and OTUs were captured by all sequencing strategies used (Tables\ \@ref(tab:venn-table-asv) and \@ref(tab:venn-table-otu)).
The  short amplicon ASVs shared between all sequencing technologies represented `r text_list(percent(colSums(mutate_all(readd(venn_ASV)[c("0111", "1111"), c("reads_frac_pb_483", "reads_frac_SH-2257", "reads_frac_is_057")], as.numeric), na.rm = TRUE)))`% <!-- make this match--> of the reads for PacBio, Illumina, and Ion Torrent, respectively (Figure\ \@ref(fig:venn)a).
When differences at the intra-species scale were removed by clustering the ASVs into 97% OTUs, the number of OTUs shared between all three technologies increased to `r sum(readd(venn_OTU)[c("0111", "1111"),"OTUs"])`, representing `r text_list(percent(colSums(mutate_all(readd(venn_OTU)[c("0111", "1111"), c("reads_frac_pb_483", "reads_frac_SH-2257", "reads_frac_is_057")], as.numeric), na.rm = TRUE)))`% of reads, respectively (Figure\ \@ref(fig:venn)b).
In particular, the majority of the `r sum(readd(venn_ASV)[c("1001","0001"),"ASVs"])` unique Ion Torrent ASVs, were found to be shared with other sequencing technologies upon OTU clustering.
ASVs unique to the Ion Torrent dataset made up `r percent(sum(as.numeric(readd(venn_ASV)[c("1001","0001"),"reads_frac_is_057"])))` of reads in that dataset, but only `r percent(sum(as.numeric(readd(venn_OTU)[c("1001","0001"),"reads_frac_is_057"])))` of belonged to a unique OTU after clustering.
In contrast, `r percent(readd(venn_ASV)["1000", "reads_frac_pb_500"])` of reads in the long PacBio dataset belonged to ASVs whose ITS2 region was unique to that dataset (Figure\ \@ref(fig:venn)c), and the fraction only reduced to `r percent(readd(venn_OTU)["1000", "reads_frac_pb_500"])` after clustering the ITS2 regions into OTUs (Figure\ \@ref(fig:venn)d).

@@

```{r accum-ASV, fig.cap="ASV accumulation curves"}
readd(accum_plot_ASV)
```

```{r accum-OTU, fig.cap="OTU accumulation curves"}
readd(accum_plot_OTU)
```

@@

For all sequencing datasets and taxonomic assignment protocols, a higher proportion of reads was assigned than of ASVs, indicating that common ASVs were more likely to be taxonomically identified than rare ASVs (Figure\ \@ref(fig:tax-chart)).
A greater fraction of ITS reads and ASVs were assigned using the Unite database than the Warcup database across sequencing technologies, amplicons, algorithms, and taxonomic ranks.
At most taxonomic ranks, the RDPC algorithm assigned the greatest fraction of reads and ASVs, followed by SINTAX, and then IDTAXA.
